---
title: "Using R to Compare the Exponential Distribution with the Central Limit Theorem"
author: "Alexander Pyle, <apyle@github.com>"
date: "December 27, 2015"
output:
  html_document:
    keep_md: yes
  pdf_document: default
---

 A sample set of headings that could be used to guide the creation of your report might be:

    (DONE) Title (give an appropriate title) and Author Name
    (DONE) Overview: In a few (2-3) sentences explain what is going to be reported on.
    (DONE) Simulations: Include English explanations of the simulations you ran, with the accompanying R code. Your explanations should make clear what the R code accomplishes.
    (DONE) Sample Mean versus Theoretical Mean: Include figures with titles. In the figures, highlight the means you are comparing. Include text that explains the figures and what is shown on them, and provides appropriate numbers.
    Sample Variance versus Theoretical Variance: Include figures (output from R) with titles. Highlight the variances you are comparing. Include text that explains your understanding of the differences of the variances.
    Distribution: Via figures and text, explain how one can tell the distribution is approximately normal. 
    
## Overview

```{r, echo=FALSE}

# load required libraries
library(data.table)
library(ggplot2)
library(xtable)

# variables used in the text and code, reinitiallized in the next chunk. Set here for use in text
lambda <- 0.2   # given by assignment
samples <- 40   # number of samples to use for the mean
simulations <- 1000     # number of simulations to run
popmean <- 1/lambda     # theoretical mean of the population
popvariance <- (1/lambda)^2 / samples # theoretical variance of the population

```

The Central Limit Theorem (CLT) posits that the distribution of averages of 
independent and identically distributed (iid) variables randomly drawn from a 
population becomes that of a population normal as the sample size increases. This 
analysis tests this theorem and determines that it holds true for mean and the 
variance. That is to say, the sample mean will converge to the population mean 
as the number of simulations of iid varables averaged together increases, and 
that the variance of these averages will converge to the population variance.

## Simulations

To demonstrate the Central Limit Theorem (CLT) this analysis will draw `r samples` 
independent and identically distributed (iid) sample values from the exponential 
distribution. These values will be averaged together. By collecting a large 
number of these averages we can observe the mean of the averages is close to the
theoretical mean of an exponential distribution population.

The theoretical mean for the exponential distribution is $\frac{1}{\lambda}$ 
while the standard deviation is also $\frac{1}{\lambda}$. Since $\lambda$ is 
`r lambda`, the mean for our population would be $\frac{1}{0.2}$ or `r popmean`. 

The following code generates `r simulations` simulations of `r samples` variables 
and stores the means of each simulation in `meanexp`. This set of simulations are 
then plotted with the `data.frame meanDF` variable.

```{r, fig.height=6, fig.width=8, fig.align='center'}

set.seed(90125) # seed the random number generator for reproducible results

lambda <- 0.2   # given by assignment
samples <- 40   # number of samples to use for the mean
simulations <- 1000     # number of simulations to run
popmean <- 1/lambda     # theoretical mean of the population
popvariance <- (1/lambda)^2 / samples # theoretical variance of the population

meanexp <- NULL
for (i in 1 : simulations) 
        meanexp <- c(meanexp, mean(rexp(samples, lambda)))

samplemean <- mean(meanexp)
meanDF <- as.data.frame(meanexp)

labYPos = 0.55  # standardize location for labels

g <- ggplot(data = meanDF, aes(x = meanexp)) 
g <- g + geom_histogram(aes(y = ..density..), fill = "lightblue", 
                        binwidth = 0.2, colour = "black")
g <- g + geom_density(size = 2, colour = "red")
g <- g + labs(title = paste("Figure 1: Distribution of Averages of", samples, "Samples"), 
              x = paste("Mean of", samples, "Samples"),
              y = "Density")
g <- g + geom_vline(x = samplemean, size = 1, colour = "black")
g <- g + annotate("text", x = samplemean - 1, y = labYPos, 
                  label = paste("Sample Mean (", round(samplemean, 3), ") -->", sep = ""))
g <- g + geom_vline(x = popmean, size = 1, colour = "black")
g <- g + annotate("text", x = popmean + 1, y = labYPos, 
                  label = paste("<-- Theoretical Mean (", round(popmean, 3), ")", sep = ""))
g

```

As we can observe from the plot above, the theoretical mean of `r round(popmean, 3)` is 
closely simulated with the sample mean of `r round(samplemean, 3)`. The density function, 
drawn in red on the plot, closely resembles a normal distribution which we would 
expect from taking the average of a draw of iid values. It does not completely 
match the normal Gaussian distribution because we did not draw an infinite number 
of values, but it already come very close simply with `r simulations` simulations. 
By simply using `r simulations` simulations of `r samples` we have quickly 
demonstrated the value of using the sample mean as a good approximation of the 
population mean the iid values have been drawn from.

## Variance

The variance for the exponential distribution is $\frac{\sigma^2}{n}$ where $n$ 
is given as `r samples`. The variance then works out to be 
$\frac{(\frac{1}{\lambda})^2}{n} = \frac{(\frac{1}{0.2})^2}{40} = \frac{5^2}{40} = \frac{25}{40} = 0.625$.


```{r}

mean100 <- NULL
for (i in 1: 100)
        mean100 <- c(mean100, mean(rexp(samples, lambda)))
mean10K <- NULL
for (i in 1:10000)
        mean10K <- c(mean10K, mean(rexp(samples, lambda)))
mean100K <- NULL
for (i in 1:100000)
        mean100K <- c(mean100K, mean(rexp(samples, lambda)))

df <- data.frame(Samples = c("100", "1,000", "10,000", "100,000"),
                 Variance = c(round(c(var(mean100), var(meanexp), var(mean10K), var(mean100K)), 4)), 
                 PopVariance = popvariance,
                 Difference = round(c(var(mean100) - popvariance, var(meanexp) - popvariance, 
                                      var(mean10K) - popvariance, var(mean100K) - popvariance), 3)
                 )

tab <- xtable(df, "html", digits = 3)
print.xtable(tab, "html", include.rownames = FALSE)


# library(ggplot2)
# nosim <- 10000;
# dat <- data.frame(
#     x = c(apply(matrix(rnorm(nosim * 40), nosim), 1, var),
#           apply(matrix(rnorm(nosim * 40), nosim * 10), 1, var),
#           apply(matrix(rnorm(nosim * 40), nosim * 100), 1, var)),
#     n = factor(rep(c("1000", "10000", "100000"), c(nosim, nosim * 10, nosim * 100)))
# )
# ggplot(dat, aes(x = x, fill = n)) + 
#         geom_density(size = 2, alpha = .2) + 
#         geom_vline(xintercept = 1, size = 2)

#The theoretical mean of exponential distribution is 1/λ and the standard deviation is also 1/λ.
#Given the Central Limit Theorem, your expected mean is
#1/lambda=1/0.2=5
#and your expected variance is
#sigma2n=(1/lambda)2n=(1/0.2)240=0.625

```

## Distribution

Refering back to Figure 1 above, we can observe the distribution density in the red line which smooths out the distribution given by the boxes in the plot. 

## Appendix

```{r, warning=FALSE}

library(devtools)
devtools::session_info() # display environment the script was create and run in.

```

```{r}

#More potentially useful code:
#
#---
#title: "Testing chunk reuse"
#author: "Neil Saunders"
#date: "24/02/2015"
#output: html_document
#---
# 
### Introduction
#Here is my very interesting document.
# 
#Chunk 1 is calling chunk 2 here, but you can't see it.
#```{r chunk1, ref.label="chunk2", echo=FALSE}
#```
# 
### This chunk is unnamed but can now use code from chunk 2
#```{r}
#myFunction(7)
#```
# 
### This is chunk 2
#My long and ugly R function is now down here.
# 
#```{r chunk2}
## it's not really long and ugly
## it just squares the input
## but imagine that it is long and ugly
# 
#myFunction <- function(x) {
#  print(x ^ 2)
#}
#```
```

